---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version-last-release/DynForest)](https://CRAN.R-project.org/package=DynForest)
[![Downloads](https://cranlogs.r-pkg.org/badges/DynForest?color=blue)](https://www.r-pkg.org/pkg/DynForest)
[![R build status](https://github.com/anthonydevaux/DynForest/workflows/R-CMD-check/badge.svg)](https://github.com/anthonydevaux/DynForest/actions)
<!-- badges: end -->

## Description

`DynForest` is a R package aiming to predict an outcome using multivariate longitudinal predictors. The method is based on random forest principle where the longitudinal predictors are modeled through the random forest. `DynForest` currently supports continuous, categorical and survival outcome. The methodology is fully described for a survival outcome in the paper:

> Devaux A., Helmer C., Genuer R. and Proust-Lima C. (2023). Random survival forests with multivariate longitudinal endogenous covariates. _Statistical Methods in Medical Research_. [<doi:10.1177/09622802231206477>](https://doi.org/10.1177/09622802231206477)

`DynForest` user guide is also available in the paper: 

> Devaux A., Proust-Lima C. and Genuer R. (2023). Random Forests for time-fixed and time-dependent predictors: The DynForest R package. _arXiv_.  [<doi:10.48550/arXiv.2302.02670>](https://doi.org/10.48550/arXiv.2302.02670)

## Installation

`DynForest` package version 1.1.2 could be install from the [CRAN](https://cran.r-project.org/package=DynForest) with:

```{r, eval = FALSE}
install.packages("DynForest")
```

Development version of `DynForest` is also available from [GitHub](https://github.com/anthonydevaux/DynForest) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("anthonydevaux/DynForest")
```

## Quick example

### Manage data

```{r}
library(DynForest)
data(pbc2)

# Get Gaussian distribution for longitudinal predictors
pbc2$serBilir <- log(pbc2$serBilir)
pbc2$SGOT <- log(pbc2$SGOT)
pbc2$albumin <- log(pbc2$albumin)
pbc2$alkaline <- log(pbc2$alkaline)
```

### Build DynForest objects

```{r}
# Build longitudinal data
timeData <- pbc2[,c("id","time",
                    "serBilir","SGOT",
                    "albumin","alkaline")]

# Create object with longitudinal association for each predictor
timeVarModel <- list(serBilir = list(fixed = serBilir ~ time,
                                     random = ~ time),
                     SGOT = list(fixed = SGOT ~ time + I(time^2),
                                 random = ~ time + I(time^2)),
                     albumin = list(fixed = albumin ~ time,
                                    random = ~ time),
                     alkaline = list(fixed = alkaline ~ time,
                                     random = ~ time))
# Build fixed data
fixedData <- unique(pbc2[,c("id","age","drug","sex")])

# Build outcome data
Y <- list(type = "surv",
          Y = unique(pbc2[,c("id","years","event")]))
```

### Run `DynForest()` function

```{r}
# Run DynForest function
res_dyn <- DynForest(timeData = timeData, fixedData = fixedData,
                     timeVar = "time", idVar = "id",
                     timeVarModel = timeVarModel, Y = Y,
                     ntree = 50, nodesize = 5, minsplit = 5,
                     cause = 2, ncores = 2, seed = 1234)
```

### Get summary

```{r}
summary(res_dyn)
```
