---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/DynForest)](https://CRAN.R-project.org/package=DynForest)
[![R-CMD-check](https://github.com/anthonydevaux/DynForest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/anthonydevaux/DynForest/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Description

`DynForest` is a R package aiming to predict an outcome using multivariate longitudinal predictors. The method is based on random forest principle where the longitudinal predictors are modeled through the random forest. `DynForest` currently supports continuous, discrete and survival outcome. The methodology is fully described for a survival outcome in the paper:

> Devaux A., Helmer C., Dufouil C., Genuer R., Proust-Lima C. (2022). Random survival forests for competing risks with multivariate longitudinal endogenous covariates. _arXiv_. [<doi: 10.48550/arXiv.2208.05801>](https://doi.org/10.48550/arXiv.2208.05801)

## Installation

`DynForest` package could be install from the [CRAN](https://cran.r-project.org/package=DynForest) with:

```{r, eval = FALSE}
install.packages("DynForest")
```

Development version of DynForest is also available from [GitHub](https://github.com/anthonydevaux/DynForest) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("anthonydevaux/DynForest")
```

## Acknowlegments

We thank Louis Capitaine for his precious help for the R code.
